---
title: "Practical Machine Learning Assignment"
output: html_document
date: "2024-10-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R Markdown
This report demonstrates building and evaluating machine learning models using Decision Tree and Random Forests in R. Below are the results and visualizations from the analysis.


## Loading Required Libraries

```{r load_libraries}
library(caret)  # For data partitioning and model evaluation
library(randomForest)  # For random forest model
library(ggplot2)  # For data visualization

# File paths and URLs for data sources
train_data_path <- './data/pml-training.csv'
test_data_path  <- './data/pml-testing.csv'
train_data_url  <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test_data_url   <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'
# Creating directories if not already present
if (!dir.exists("data")) {
  dir.create("data")
}
if (!dir.exists("data/results")) {
  dir.create("data/results")
}
Setting Seed for Reproducibility
# Setting seed for reproducibility of the analysis
set.seed(15342)

```
Reading and Cleaning the Datasets
```{r}
# Reading the datasets and handling missing values
train_data <- read.csv(train_data_path, na.strings = c("NA", "#DIV/0!", ""))
test_data <- read.csv(test_data_path, na.strings = c("NA", "#DIV/0!", ""))

# Dropping columns with any missing data
train_data_clean <- train_data[, colSums(is.na(train_data)) == 0]
test_data_clean <- test_data[, colSums(is.na(test_data)) == 0]

# Removing unnecessary columns like IDs and timestamps
train_data_clean <- train_data_clean[, -c(1:7)]
test_data_clean <- test_data_clean[, -c(1:7)]



```

Splitting the Training Data
```{r}
index_train <- createDataPartition(train_data_clean$classe, p = 0.75, list = FALSE)
training_set <- train_data_clean[index_train, ]  # Ensure you use 'training_set' here
validation_set <- train_data_clean[-index_train, ]

```
## Splitting the Training Data
# Building a Decision Tree model
tree_model <- rpart(classe ~ ., data = training_set, method = "class")

# Making predictions using the Decision Tree model
tree_predictions <- predict(tree_model, validation_set, type = "class")

# Visualizing the Decision Tree structure
rpart.plot(tree_model, main = "Classification Tree", extra = 104, under = TRUE)

# Evaluating the Decision Tree model with a confusion matrix
confusionMatrix(tree_predictions, validation_set$classe)


```

```{r split_data}
index_train <- createDataPartition(train_data_clean$classe, p = 0.75, list = FALSE)
training_set <- train_data_clean[index_train, ]  # Ensure you use 'training_set' here
validation_set <- train_data_clean[-index_train, ]
# Making predictions for the final test dataset using the trained Random Forest model
final_predictions <- predict(rf_model, test_data_clean, type = "class")

# Displaying the predictions for submission
final_predictions
